---
title: 'COVID19 WikiPathways & DiseaseMap pathway and network analysis'
author: "Finterly Hu, Martina Summer-Kutmon"
date: "10 Feb 2021"
comment: "adapted from Alexander Pico's Pathway Analysis Vignette"
version: 1.0
---

# General instructions (read before running the code snippets)
* This script contains six code snippets from pre-processing to pathway analysis to pathway visualization, crosstalk and network analysis. The steps should be run one after the other. 
* Make sure all the required packages are installed beforehand (BiocManager::install(...)). 
* Make sure you have Cytoscape installed (version 3.8.0+) and running before you start running the script. 

# Setup

Loading libraries
```{r}
library(readr)
library(DESeq2)
library(here)
library(dplyr)
library(tibble)
library(clusterProfiler)
library(org.Hs.eg.db)
library(tidyverse)
library(rWikiPathways)
library(RCy3)
library(RColorBrewer)
library(EnhancedVolcano)
library(readr)
library(reshape)
library(data.table)
```

Setup
```{r}
setwd(here())
```


# Load dataset

We take the series 1 from the following dataset (NHBE mock treated versus SARS-CoV-2 infected): 
* https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147507

which is related to the following publication:

Blanco-Melo, Daniel, et al. "Imbalanced host response to SARS-CoV-2 drives development of COVID-19." Cell 181.5 (2020): 1036-1045.

```{r}
GSE147507_raw_counts <- read.csv("data/GSE147507_RawReadCounts_Human.tsv", sep = "\t")

NHBEvsCOV2.df <- GSE147507_raw_counts[,c(2:7)]
row.names(NHBEvsCOV2.df) <- GSE147507_raw_counts$X

NHBEvsCOV2.targets <- as.data.frame(matrix(NA,length(names(NHBEvsCOV2.df)),1))
names(NHBEvsCOV2.targets) <- c("condition")
row.names(NHBEvsCOV2.targets) <- names(NHBEvsCOV2.df)
NHBEvsCOV2.targets$condition <- gsub("Series1_", "", row.names(NHBEvsCOV2.targets))
NHBEvsCOV2.targets$condition <- factor(gsub("_[1-3]$", "", NHBEvsCOV2.targets$condition))

NHBEvsCOV2.dds <- DESeqDataSetFromMatrix(countData = as.matrix(NHBEvsCOV2.df),colData = NHBEvsCOV2.targets, design = ~ condition)
## Set control
NHBEvsCOV2.dds$condition <- relevel(NHBEvsCOV2.dds$condition, ref = levels(NHBEvsCOV2.targets$condition)[1])
## Carry out diff exp
NHBEvsCOV2.dds <- DESeq(NHBEvsCOV2.dds)
## See the comparisons carried out
NHBEvsCOV.comp <- resultsNames(NHBEvsCOV2.dds)
## Get results table
NHBEvsCOV.res <- results(NHBEvsCOV2.dds, name=NHBEvsCOV.comp[2])

NHBEvsCOV.res.df <- as.data.frame(NHBEvsCOV.res) %>% 
    tibble::rownames_to_column(var = "GeneID") %>% 
    dplyr::select(c(GeneID, log2FoldChange, pvalue, padj, stat)) %>% 
    tibble::column_to_rownames(var = "GeneID") 

NHBEvsCOV.res.df <- setDT(NHBEvsCOV.res.df, keep.rownames = TRUE)[]
colnames(NHBEvsCOV.res.df)[1] <- "GeneID"

svg('figures/vulcanoplot.svg')
EnhancedVolcano(NHBEvsCOV.res.df, title = "NHBE cell line", lab = NHBEvsCOV.res.df$GeneID, x = 'log2FoldChange', y = 'pvalue', pCutoff = 0.05, FCcutoff = 0.585)
dev.off()
png('figures/vulcanoplot.png')
EnhancedVolcano(NHBEvsCOV.res.df, title = "NHBE cell line", lab = NHBEvsCOV.res.df$GeneID, x = 'log2FoldChange', y = 'pvalue', pCutoff = 0.05, FCcutoff = 0.585)
dev.off()

# identifier mapping 
hgcn2entrez <- clusterProfiler::bitr(NHBEvsCOV.res.df$GeneID, fromType = "SYMBOL",toType = c("ENTREZID","SYMBOL","ENSEMBL"), OrgDb = org.Hs.eg.db)
data <- merge(NHBEvsCOV.res.df, hgcn2entrez, by.x="GeneID", by.y="SYMBOL", all.x = TRUE)
data <- subset(data, select=c(6,1:5,7))

# filter genes without Entrez Gene identifier
data.NHBE <- data %>% tidyr::drop_na(ENTREZID)
colnames(data.NHBE)[2] <- "GeneName"
colnames(data.NHBE)[1] <- "GeneId"

write.table(data.NHBE, file="output/DEG_Data.txt", sep="\t", quote=FALSE, row.names = FALSE)
```


# Pathway enrichment

Pathway enrichment is performed with the gene sets of WikiPathways and the COVID-19 Disease Map maps.

```{r}
deg <- unique(data.NHBE[!is.na(data.NHBE$pvalue) & data.NHBE$pvalue < 0.05 & abs(data.NHBE$log2FoldChange) > 0.58,c(1,2)])

bkgd.genes <- unique(data.NHBE[,c(1,2)])

# combine WikiPathways and COVID19 Disease Map gene sets
wp2gene <- readPathwayGMT("data/wikipathways-20210110-gmt-Homo_sapiens.gmt")
dm2gene <- readPathwayGMT("data/COVID19_DiseaseMap_April2021.gmt")
pwy2gene <- dplyr::bind_rows(wp2gene, dm2gene)

wpid2gene <- pwy2gene %>% dplyr::select(wpid,gene) #TERM2GENE
wpid2name <- pwy2gene %>% dplyr::select(wpid,name) #TERM2NAME

ewp <- clusterProfiler::enricher(
  deg$GeneId,
  universe = bkgd.genes$GeneId,
  pAdjustMethod = "fdr",
  pvalueCutoff = 0.05,
  TERM2GENE = wpid2gene,
  TERM2NAME = wpid2name)

ewp.res <- as.data.frame(ewp) 

# export enrichment result
svg('figures/barplot.svg', width = 800, height=1000)
barplot(ewp, showCategory = 45)
dev.off()
png('figures/barplot.png', width = 800, height=1000)
barplot(ewp, showCategory = 45)
dev.off()

write.table(ewp.res, file="output/enrich_res.txt", sep="\t", quote=FALSE, row.names = FALSE)
```


# Pathway data visualization

The pathways can then be visualized with the gene expression data as shown with the 
"Type I Interferon Induction and Signaling During SARS-CoV-2 Infection" (WP4868) pathway from the WikiPathways COVID-19 portal. 

```{r}
RCy3::commandsRun('wikipathways import-as-pathway id=WP4868') 
toggleGraphicsDetails()
loadTableData(data.NHBE, data.key.column = "ENSEMBL", table.key.column = "Ensembl")

# apply visual style 
data.values = c(-1,0,1) 
node.colors <- c(rev(brewer.pal(length(data.values), "RdBu")))
setNodeColorMapping("log2FoldChange", data.values, node.colors, default.color = "#D8D8D8", style.name = "WikiPathways")

RCy3::setVisualStyle("WikiPathways")

# Saving output
svg.file <- file.path(getwd(), "figures/PathwayVisualization.svg")
exportImage(svg.file,'SVG')
png.file <- file.path(getwd(), "figures/PathwayVisualization.png")
exportImage(png.file,'PNG', zoom = 500)
cys.file <- file.path(getwd(), "output/PathwayVisualization.cys")
saveSession(cys.file) 

#comment following line if you want to manipulate the visualization in Cytoscape
RCy3::closeSession(save.before.closing = F)
```


# Pathway overlap visualization

There is often crosstalk and overlap between pathways enriched in gene expression analyses. The following step visualizes the overlap between the enriched pathways in a pathway-gene network. 

The genes not present in any pathway are included in the visualization but can be filtered in a follow-up step if preferred. 

```{r}
pwy <- unique(ewp.res[,c(1,2)])
colnames(pwy) <- c("id","label")
pwy$type <- 'pathway'

edges <- wpid2gene[wpid2gene$wpid %in% pwy$id,]
colnames(edges) <- c("source", "target")

genes <- unique(deg)
colnames(genes) <- c("id","label")
genes$type <- 'gene'

edges <- unique(edges[edges$target %in% genes$id,])

nodes <- dplyr::bind_rows(genes, pwy)
rownames(nodes) <- NULL

createNetworkFromDataFrames(nodes=nodes,edges=edges,title="Pathway-Gene-Associations", collection="PathwayGeneCrosstalk")

loadTableData(data.NHBE, data.key.column = "GeneId", table.key.column = "id")

# Visual style
RCy3::copyVisualStyle("default","wp.vis")

RCy3::setNodeLabelMapping("label", style.name="wp.vis")
RCy3::lockNodeDimensions(TRUE, style.name="wp.vis")
RCy3::setNodeShapeMapping('type', c('gene','pathway'), c("ellipse","hexagon"), style.name="wp.vis")
RCy3::setNodeSizeMapping('type', c('gene','pathway'), c(40,25), mapping.type = "d", style.name = "wp.vis")
data.values<-c(-1,0,1) 
node.colors <- c(rev(brewer.pal(length(data.values), "RdBu")))
setNodeColorMapping("log2FoldChange", data.values, node.colors, default.color = "#99FF99", style.name = "wp.vis")

RCy3::setVisualStyle("wp.vis")

# Saving output
svg.file <- file.path(getwd(), "figures/PathwayCrosstalk.svg")
exportImage(svg.file,'SVG')
png.file <- file.path(getwd(), "figures/PathwayCrosstalk.png")
exportImage(png.file,'PNG', zoom = 500)
cys.file <- file.path(getwd(), "output/PathwayCrosstalk.cys")
saveSession(cys.file) 

#comment following line if you want to manipulate the visualization in Cytoscape
RCy3::closeSession(save.before.closing = F)
```


# Drug-target extension
Drug-target information is extracted from DrugBank
Wishart DS, Knox C, Guo AC, Shrivastava S, Hassanali M, Stothard P, Chang Z, Woolsey J. Drugbank: a comprehensive resource for in silico drug discovery and exploration. Nucleic Acids Res. 2006 Jan 1;34 (Database issue):D668-72. 16381955.

Please check the DrugBank license information
https://go.drugbank.com/legal/terms_of_use

License: 
https://creativecommons.org/licenses/by-nc/4.0/legalcode

```{r}
pwy <- unique(ewp.res[,c(1,2)])
colnames(pwy) <- c("id","label")
pwy$type <- 'pathway'

edges <- wpid2gene[wpid2gene$wpid %in% pwy$id,]
colnames(edges) <- c("source", "target")

genes <- unique(deg)
colnames(genes) <- c("id","label")
genes$type <- 'gene'

edges <- edges[edges$target %in% genes$id,]

nodes <- dplyr::bind_rows(genes, pwy)
rownames(nodes) <- NULL

createNetworkFromDataFrames(nodes=nodes,edges=edges,title="Gene-drug interactions", collection="drug-target")

drugbank <- file.path(getwd(), "data/drugbank-5.1.7.xgmml")


# run CyTargetLinker
commandsRun(paste0('cytargetlinker extend idAttribute="id" linkSetFiles="', drugbank, '"'))
commandsRun('cytargetlinker applyLayout network="current"')

drug <- read.delim2("data/drugNodes-5.1.7.txt", stringsAsFactors=TRUE)
loadTableData(drug, data.key.column="Identifier", table.key.column="id")
loadTableData(data.NHBE, data.key.column = "GeneId", table.key.column = "id")

# Visual style
RCy3::copyVisualStyle("default","wp.vis")
RCy3::setNodeLabelMapping("label", style.name="wp.vis")
RCy3::lockNodeDimensions(TRUE, style.name="wp.vis")
RCy3::setNodeShapeMapping('type', c('gene','pathway','drug'), c("ellipse","rectangle","hexagon"), style.name="wp.vis")
RCy3::setNodeSizeMapping('type', c('gene','pathway'), c(40,25), mapping.type = "d", style.name = "wp.vis")
data.values = c(-1,0,1) 
node.colors <- c(rev(brewer.pal(length(data.values), "RdBu")))
data.values<-c(-1,0,1) 
setNodeColorMapping("log2FoldChange", data.values, node.colors, default.color = "#99FF99", style.name = "wp.vis")

RCy3::setVisualStyle("wp.vis")

my.drugs <- selectNodes("drug", by.col = "CTL.Type", preserve = FALSE)$nodes 
clearSelection()
setNodeColorBypass(my.drugs, "#DD99FF")

# Saving output
svg.file <- file.path(getwd(), "figures/DrugExtension.svg")
exportImage(svg.file,'SVG')
png.file <- file.path(getwd(), "figures/DrugExtension.png")
exportImage(png.file,'PNG', zoom = 500)
cys.file <- file.path(getwd(), "output/DrugExtension.cys")
saveSession(cys.file) 

#comment following line if you want to manipulate the visualization in Cytoscape
RCy3::closeSession(save.before.closing = F)
```

# PPI network analysis - MINERVA pathways (COVID 19 Disease Map)

```{r}
x <- format_csv(as.data.frame(deg$GeneId), col_names=F, quote_escape = "double", eol =",")
commandsRun(paste0('string protein query cutoff=0.9 newNetName="PPI network" query="',x,'" limit=0'))

# Saving output
svg.file <- file.path(getwd(), "figures/PPI-Network.svg")
exportImage(svg.file,'SVG')
png.file <- file.path(getwd(), "figures/PPI-Network.png")
exportImage(png.file,'PNG', zoom = 500)

proteins <- RCy3::getTableColumns(columns=c("query term", "display name"))
ppis <- RCy3::getTableColumns(table="edge", columns=c("name"))
ppis <- data.frame(do.call('rbind', strsplit(as.character(ppis$name),' (pp) ',fixed=TRUE)))
ppis.2 <- merge(ppis, proteins, by.x="X1", by.y="display name", all.x=T)
colnames(ppis.2) <- c("s", "t", "source")
ppis.3 <- merge(ppis.2, proteins, by.x="t", by.y="display name", all.x=T)
colnames(ppis.3)[4] <-"target"
ppis.3$interaction <- "PPI"

colnames(proteins) <- c("id","label")
proteins$type <- "protein"

pwy.covid <- rWikiPathways::getPathwayIdsByCurationTag("Curation:COVID19")
pwy2gene.minerva <- pwy2gene[(grepl("MINERVA", pwy2gene$name) | pwy2gene$wpid %in% pwy.covid) & pwy2gene$gene %in% deg$GeneId,]
colnames(pwy2gene.minerva)[3] <- c("source")
colnames(pwy2gene.minerva)[5] <- c("target")
pwy2gene.minerva$interaction <- "Pathway-Gene"

pwy.filtered <- unique(pwy2gene[(grepl("MINERVA", pwy2gene$name) | pwy2gene$wpid %in% pwy.covid),c(1,3)])
colnames(pwy.filtered) <- c("label", "id")
pwy.filtered$type <- "pathway"

nodes.ppi <- dplyr::bind_rows(genes,pwy.filtered)
rownames(nodes.ppi) <- NULL
edges.ppi <- unique(dplyr::bind_rows(ppis.3[,c(3,4,5)], pwy2gene.minerva[,c(3,5,6)]))
rownames(edges.ppi) <- NULL

RCy3::createNetworkFromDataFrames(nodes= nodes.ppi, edges = edges.ppi, title="PPI-Pathway-Network2", collection="new")
loadTableData(nodes.ppi, data.key.column = "id", table="node", table.key.column = "id")
loadTableData(data.NHBE, data.key.column = "GeneId", table.key.column = "id")

# Visual style
RCy3::copyVisualStyle("default","wp.vis")
RCy3::setNodeLabelMapping("label", style.name="wp.vis")
RCy3::lockNodeDimensions(TRUE, style.name="wp.vis")
RCy3::setNodeShapeMapping('type', c('gene','pathway','drug'), c("ellipse","rectangle","hexagon"), style.name="wp.vis")
RCy3::setNodeSizeMapping('type', c('gene','pathway'), c(40,25), mapping.type = "d", style.name = "wp.vis")
node.colors <- c(rev(brewer.pal(length(data.values), "RdBu")))
data.values = c(-1,0,1) 
setNodeColorMapping("log2FoldChange", data.values, node.colors, default.color = "#99FF99", style.name = "wp.vis")

RCy3::setVisualStyle("wp.vis")

# Saving output
svg.file <- file.path(getwd(), "figures/PPI-Pathway-Network.svg")
exportImage(svg.file,'SVG')
png.file <- file.path(getwd(), "figures/PPI-Pathway-Network.png")
exportImage(png.file,'PNG', zoom = 500)
cys.file <- file.path(getwd(), "output/PPI-Pathway-Network.cys")
saveSession(cys.file) 

#comment following line if you want to manipulate the visualization in Cytoscape
RCy3::closeSession(save.before.closing = F)

```
